The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-06-17 paper/ms.tex
rewriting the paper

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider cases of a single central \lya source, and a uniform
distribution of sources.  
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis the intensity at the line
center and the line width increase with rotational velocity.   
Along the same sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux nor in the escape fraction. 
We provide an analytical expression for the outgoing line intensity
that accounts for the velocity and viewing angle trends reported here.
\end{abstract}

@@

A more detailed quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}. Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06,Ahn2014}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12, Garel2012} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies \citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

@@

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results. In
the last section we present our conclusions.

@@

The initialization process for the \ly photons especifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the restframe of the gas and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

@@

The central result of this paper are summarized in Figures
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. They show 2D histograms
of the escape frequency $x$ and outgoing angle $\theta$ parameterized by
$|\mu|$. Constructing the emission line with photons around a value
of $\mu$ would gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. We have veryfied
that the solutions are indeed symmetric with respect to $\mu=0$.

From these Figures we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  In the
next subsections we describe in  detail the changes of the morphology
with velocity, optical depth and viewing angle. We characterize the
line morphology by its total intensity, the full width at half maximum
(FWHM) and the location of the peak maxima. In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. For the models where
dust is included we measure the escape fraction as a function of
rotational velocity.

@@

If the viewing angle is aligned witht the rotation axis, $|\mu|\sim
1$, the \ly line keeps a double peak with minor changes with
increasing rotational velocity. However, for a line of sight
perpendicular to the rotation axis, $|\mu|\sim 0$, the impact of
rotation is larger. This is clear in  Figure
\ref{fig:differentvelocities} where we present the different line
morphologies for $|\mu|\sim 0$ for both the homogeneous and central
configurations. The panels have the same distribution as Figures
\ref{fig:CentralSpec} and \ref{fig:HomSpec}.

In the case of $|\mu|\sim 0$ there are three clear effects on the
line morphology as the rotational velocity increases. First, the
line broadens; second, the double peaks reduce their intensity; and
third, the intensity at the line center rises. The last two effects
merge to give the impression that the double peaks are merged into one
at high rotational velocities, a result that is evident for the
homogeneously distributed sources as shown in the dashed lines of
Figure \ref{fig:differentvelocities}.

@@

Besides the change in morphology, we now consider possible variations
in the integrated flux with respect to the viewing angle $\theta$. We
define the normalized flux seen by an observer at an angle $\mu$ by:

\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation} 
%
where $\mu=\cos\Theta$, $N$ is the total number of outgoing photons,
$\Delta N$ is the number of photons in an angular bin $\Delta
\Theta$. This definition satisfies the condition
$\int_{-1}^{1}F(\mu)d\mu/2=1$.  In the case of perfect spherical
simmetry one expects a flat distribution of $F(\mu)=1$.

Figure \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and both
for the central and homogeneous source distributions. The
distributions do not show any trend, beyond some fluctuations below
the $2\%$ level.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f7.pdf}
\end{center}
\caption{FWHM for the non-dusty models as a function of
  rotational velocity $V_{\rm max}$. The left panel shows the
  results in velocity units while the right panel normalizes
  the data by the FWHM in the static case. 
  Continuous (dashed) lines  correspond to central (homogeneous)
  source distributions. The straight lines represent the fit to
  the data using the expression in Eq. (\ref{eq:fwhm}).
  \label{fig:widthsvsvelocity}}
\end{figure*}

We use the full width at half maximum (FWHM) to quantify the line
broadening. We measure this width from the velocity histogram by
finding the values of the velocities at half maximum intensity. We use
lineal interpolation between histogram points to get a value more precise
than the bin size used to construct the line intensity.

Figure \ref{fig:widthvsmu} shows the FWHM for all models as a function
of the viewing angle. The FWHM increases for decreasing values of
$\mu$ and increasing values of $V_{\rm max}$. In Figure
\ref{fig:widthvsvelocity} we fix $|\mu|<0.1$, i.e. viewing angle
perpendicular to the rotation axis, and plot the FWHM as a
function of rotational velocity.

we parametrize the dependency of the line width with  $V_{\rm max}$ as
%
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the velocity width in the static case and $\lambda$ 
is a positive scalar to be determined as a fit to the data.  With this
test we want to know to what extent the new velocity width can be
expressed as a quadratic sum of the two relevant velocities in the
problem. That all the models fall into two families of straight lines,
for central and homogeneously emitted sources, justifies this choice
of parameterization.

We fit simultaneously all the points corresponding to each one of this
families to find $\lambda_{\rm c} = 0.84 \pm 0.06$ and $\lambda_{\rm
  h}= 0.54\pm 0.10$ respectively.

@@

\appendix
\section{Analytic Expression for Ly$\alpha$ Spectrum
  emerging from Rotating Cloud}

Ly$\alpha$ scattering through a an optically thick gas cloud that is
undergoing solid-body rotation (i.e. the angular speed around the
rotation axis is identical for each hydrogen atom) proceeds identical
as in a static cloud. While scattering events off atoms within the
rotating cloud impart Doppler boosts on the Ly$\alpha$ photon, these
Doppler boost are only there in the lab-frame. In the frame of the
rotating gas cloud, all atoms are stationary w.r.t each other, and the
scattering process proceeds identical as in the static
case\footnote{An alternative - more quantitative way - to view this is
  to consider the path of an individual photon: say the photon travels
  a distance corresponding to $\tau=1$, scatters by 90 degrees, then
  travels another distance corresponding to $\tau=1$. It is easy to
  show that the point where the photon ends up is independent of the
  assumed rotation velocity of the cloud.}. This explains why the
escape fraction of Ly$\alpha$ photon is independent of the rotation
velocity off the cloud, irrespective of the rotation speed.

The spectrum emerging of photons emerging from a rotating gas cloud is
also identical as for the static case in a frame that is co-rotating
with the cloud. However, the surface of cloud now moves in the
lab-frame. Each surface-element on the rotating cloud now has a bulk
velocity w.r.t a distant observer. In order to compute the spectrum
emerging from this rotation cloud, we have to sum over all surface
elements (weighted by their intensity).

To provide an analytic description of this emerging spectrum, we adopt
the geometry shown in Fig~1 (note this differs slightly from the main
body of the paper. My `observer'  sits along the x-axis. I think yours
sits along the z-axis. We have to makes sure this is done consistently
throughout the paper).  
\begin{figure*}[h]
\centerline{\includegraphics[width=85mm]{scheme1.pdf}
\includegraphics[width=85mm]{scheme2.pdf}}
\caption[]{Adopted geometry for evaluating the analytic spectrum.}
\label{fig:scheme}
\end{figure*} 
The sightline to the observer \& rotation define the $x-z$ plane. The observer sits along the axis. The {\it left panel} in Fig~1 shows the view from the $y-axis$. The rotation axis makes an angle $i$ with respect to the z-axis. We sum up spectra from individual patches by integrating over the impact parameter $b$, and angle $\phi$. Each $(b,\phi)$ corresponds to a point on the sphere. This point has a velocity vector ${\bf v}(b,\phi,i)$, which we denote with ${\bf v}$ for brevity.

\begin{displaymath}
J(x,i)=2\pi \int_0^Rdb \hs b \int_0^{2\pi}d\phi \hs
S(b,\phi)J(x,b,\phi,i) \approx 2\pi \int_0^Rdb \hs b
\int_0^{2\pi}d\phi \hs J(x,b,\phi,i)\\ \nonumber
\end{displaymath}

\begin{displaymath}
J(x,b,\phi,i)=\frac{\sqrt{\pi}}{\sqrt{24}a\tau_0}\Bigg{(}\frac{(x-x_{\rm b})^2}{1+{\rm cosh}\Big{[}\sqrt{\frac{2\pi^3}{27}}\frac{|(x -x_{\rm b})^3|}{a\tau_0}\Big{]}}\Bigg{)} 
\end{displaymath}

The velocity $v_b$ is the component of ${\bf v}$ projected onto the line-of-sight. This component is given by
\begin{equation}
v_{\rm b}(b,\phi,i)=v_{\rm rot}\sqrt{1-(b/R)^2\cos^2 \phi}\hs \cos i
\hs \sin \beta 
\end{equation}

\begin{equation}
\tan \beta= \frac{ b\sin \phi}{\sqrt{R^2 -b^2}}.
\end{equation}
%
The quantity $S(b,\phi)$ denotes the surface brightness of the sphere at $(b,\phi)$. We have assumed in our analytic calculations that $S(b,\phi)$ is constant. The assumption $S(b,\phi)=$ constant corresponds to $I(\mu) \propto \mu$ at the surface, where $\mu$ denotes the cosine of the angle of the propagation direction of the outgoing photon and the normal to the spheres surface: a fixed $db$ corresponds to a physical length $ds = db/\mu$ on the sphere. If $I(\mu)$ were constant, this would imply that the sphere should appear brighter per unit $b$. A constant surface brightness profile requires the directional dependence for $I(\mu) \propto \mu$ to correct for this. Indeed, this is what is expected for the escape of Ly$\alpha$ photons from static, extremely opaque media (see Ahn, Lee \& Lee, 2001; their Fig~4 and accompanying discussion).

It is worth stressing that the analytic calculation is clearly not a complete analytic calculation, and we do not expect perfect agreement: we {\it assumed} a functional form for the surface brightness profile [or for $I(\mu)$]. It is important to realise that $I(\mu)$ itself may depend on the spectrum. I.e., analytic solutions exist for $J(x) = \int_0^1 I(x,\mu) d\mu$ at the boundary, and {\it approximate} expressions for $I(\mu) =\int dx I(x,\mu)$, but none for $I(x,\mu)$ itself.  \\

Fig~2 shows some examples of analytic vs. full MC spectra. The area under all spectra are normalised to $1$. The MC simulations used $10^5$ photons. The two methods clearly give good agreement, though not perfect. In particular, Fig~3 shows that the MC gives rise to a spectrum that is slightly more concentrated to the centre. It is easy to get better agreement my modifying the surface brightness profile (e.g. if $S(b)=2-b$, then the analytic calculation agrees better) In any case, the analytic calculation closely captures the results obtained from the full calculations from the MC simulations. As such, they are extremely useful \& provide us with a quick tool to verify our calculations at the first order level.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{vary_vel.pdf}
  \includegraphics[width=0.48\textwidth]{vary_angle.pdf}
\end{center}
\caption{
Comparison of the Monte Carlo results agains the analytic solution.
\label{fig:comparison} }   
\end{figure*}

>>>>

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions: single central and a
homogeneoulsy distributed in the sphere. 
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis the intensity at the line
center and the line width increase with rotational velocity.   
Along the same sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux nor in the escape fraction. 
We provide an analytical expression for the outgoing \lya spectrum of
a rotating cloud which accounts for the velocity and viewing angle
trends observed in the simulations.
\end{abstract}

@@

A more detailed quantitative description of the \ly line has been
reached through Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. 
In the last two decades these studies have become popular due to the
availability of computing power. 
Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}. 
Later on, the effects of clumpy media \citep{Hansen06} and
expanding/contracting shell/spherical geometries started to be studied
\citep{Ahn03,Verhamme06,Dijkstra06}.  
Similar codes have applied these results to semi-analytic models of
galaxy formation \citep{Orsi12, Garel2012} and results of large
hydrodynamic simulations \citep{CLARA,Forero12,Behrens13}. 
Recently, Monte Carlo codes have also been applied to the results of
high resolution hydrodynamic simulations of individual galaxies
\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. 
Meanwhile, recent  developments have been focused on the systematic
study of clumpy outflows \citep{DijkstraKramer} and anisotropic
velocity configurations \citep{Zheng2013}.

@@

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results. In
the last section we present our conclusions. The Appendix presents the
derivation of an analytic expression that can be helpful to interpret
to first order the main trends observed in the Monte Carlo simulations.

@@

The initialization process for the \ly photons especifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the gas reference frame and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

@@

The central result of this paper are summarized in Figures
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parameterized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have veryfied that the solutions are indeed symmetric with respect
to $\mu=0$.

From these Figures we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximu, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity.

@@

If the viewing angle is aligned witht the rotation axis, $|\mu|\sim
1$, the \ly line keeps in most of the cases a double peak with minor
changes as the rotational velocity starts to increase. 
However, for a line of sight perpendicular to the rotation axis,
$|\mu|\sim 0$, the impact of  rotation is larger.  
The double peak readily transforms into a single peak.

This is clear in  Figure \ref{fig:differentvelocities} where we
present the different line morphologies for $|\mu|\sim 0$ in the
homogeneous and central configurations.  
The panels have the same distribution as Figures \ref{fig:CentralSpec}
and \ref{fig:HomSpec}.    
There are three clear effects on the line morphology as the rotational
velocity increases.  
First, the line broadens; second, the double peaks reduce their intensity; and
third, the intensity at the line centre rises. 
The last two effects are combined to give the impression that the double
peaks are merged into a single one at high rotational velocities.

@@

We now consider possible variations in the integrated flux with
respect to the viewing angle $\theta$. 
To this end we  define the normalized flux seen by an observer at an
angle $\mu$ by:

\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation} 
%
where $\mu=\cos\theta$, $N$ is the total number of outgoing photons,
$\Delta N$ is the number of photons in an angular bin $\Delta
\Theta$. This definition satisfies the condition
$\int_{-1}^{1}F(\mu)d\mu/2=1$.  In the case of perfect spherical
simmetry one expects a flat distribution with $F(\mu)=1$.

Figure \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, ifferent rotational velocities and the two
types of source distributions.
This Figure shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of \%2.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f7.pdf}
\end{center}
\caption{FWHM for the non-dusty models as a function of
  rotational velocity $V_{\rm max}$ for observers located
  perpendicular to the rotation axis. 
  The left panel shows the results in velocity units while the right
  panel normalizes the data by the FWHM in the static case. 
  Continuous (dashed) lines  correspond to central (homogeneous)
  source distributions. 
  The straight lines represent the fit to the data using the
  expression in Eq. (\ref{eq:fwhm}). 
  \label{fig:widthsvsvelocity}}
\end{figure*}

We use the full width at half maximum (FWHM) to quantify the line
broadening. 
We measure this width from the line intensity histogram by finding the
values of the velocities at half maximum intensity.
We use lineal interpolation between histogram points to get a value
more precise than the bin size used to construct the histogram.

Figure \ref{fig:widthvsmu} shows the FWHM for all models as a function
of the viewing angle. 
The FWHM increases for decreasing values of $\mu$ (movement from the
poles to the equator) and increasing values of $V_{\rm max}$. 
In Figure
\ref{fig:widthsvsvelocity} we fix $|\mu|<0.1$, i.e. viewing angle
perpendicular to the rotation axis, to plot the FWHM as a function of
rotational velocity.

Ee parametrize the dependency of the line width with  $V_{\rm max}$ as
%
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the velocity width in the static case and $\lambda$ 
is a positive scalar to be determined as a fit to the data.  
With this test we want to know to what extent the new velocity width can be
expressed as a quadratic sum of the two relevant velocities in the
problem.

All the models fall into a single familiy of lines in the plane shown
in the right panel of Figure \ref{fig:widthsvsvelocity}, justifying
the choice of our parameterization.
We fit simultaneously all the points in two separate groups, central
and homogeneous sources.
We find that these values are $\lambda_{\rm c} = 0.83 \pm 0.06$ and
$\lambda_{\rm    h}= 0.82\pm 0.05$ respectively.

@@

\appendix
\section{Analytic Expression for the Ly$\alpha$ Spectrum
  emerging from Rotating Cloud}

Ly$\alpha$ scattering through a an optically thick gas cloud that is
undergoing solid-body rotation (i.e. the angular speed around the
rotation axis is identical for each hydrogen atom) proceeds identical
as in a static cloud. 
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame. 
In the frame of the rotating gas cloud, all atoms are stationary w.r.t
each other, an the scattering process proceeds identical as in the static
case\footnote{An alternative - more quantitative way - to view this is
  to consider the path of an individual photon: say the photon travels
  a distance corresponding to $\tau=1$, scatters by $90$ degrees, then
  travels another distance corresponding to $\tau=1$. It is easy to
  show that the point where the photon ends up is independent of the
  assumed rotation velocity of the cloud.}. 
This explains why the escape fraction of Ly$\alpha$ photon is
independent of the rotation velocity off the cloud.

The spectrum of \lya photons emerging from a rotating gas cloud is
also identical as for the static case in a frame that is co-rotating
with the cloud. However, the surface of cloud now moves in the
lab-frame. Each surface-element on the rotating cloud now has a bulk
velocity w.r.t a distant observer. In order to compute the spectrum
emerging from this rotation cloud, we have to sum over all surface
elements (weighted by their intensity).

To provide an analytic description of this emerging spectrum, we adopt
the geometry shown in Fig~\ref{fig:scheme}. 
Note that this differs from the scheme in Fig~1 from the main body of
the paper.

\begin{figure*}[h]
\centerline{\includegraphics[width=85mm]{scheme1.pdf}
\includegraphics[width=85mm]{scheme2.pdf}}
\caption[]{Adopted geometry for evaluating the analytic spectrum.}
\label{fig:scheme}
\end{figure*}

The sightline to the observer \& rotation define the $x-z$ plane. 
The {\it left panel} in Fig~\ref{fig:scheme} shows the view from the $y$-axis. 
The observer sits along the $x$ axis. 
The rotation axis makes an angle $i$ with respect to the $z$-axis. 
We sum up spectra from individual patches by integrating over the
impact parameter $b$, and angle $\phi$. 
Each $(b,\phi)$ corresponds to a point on the sphere. 
This point has a velocity vector ${\bf v}(b,\phi,i)$, which we denote
with ${\bf v}$ for brevity.

The intensity at the surface point $(b,\phi)$ is

\begin{displaymath}
J(x,b,\phi,i)=\frac{\sqrt{\pi}}{\sqrt{24}a\tau_0}\Bigg{(}\frac{(x-x_{\rm
    b})^2}{1+{\rm cosh}\Big{[}\sqrt{\frac{2\pi^3}{27}}\frac{|(x
      -x_{\rm b})^3|}{a\tau_0}\Big{]}}\Bigg{)},
\end{displaymath} 
%
where $x_b\equiv v_{b}/v_{\rm th}$, and $v_b$ is the component of ${\bf v}$ projected onto the line-of-sight. This component is given by
\begin{equation}
v_{\rm b}(b,\phi,i)=V_{\rm max}\sqrt{1-(b/R)^2\cos^2 \phi}\hs \cos i
\hs \sin a,
\end{equation}
%
where

\begin{equation}
\tan a = \frac{ b\sin \phi}{\sqrt{R^2 -b^2}}.
\end{equation}

In order to compute to total intensity we integrate over $b$ and
$\phi$ with a weight wigen by the surface brightness of the
sphere at $(b,\phi)$, $S(b,\phi)$.

\begin{displaymath}
J(x,i)=2\pi \int_0^Rdb \hs b \int_0^{2\pi}d\phi \hs
S(b,\phi)J(x,b,\phi,i) \approx 2\pi \int_0^Rdb \hs b
\int_0^{2\pi}d\phi \hs J(x,b,\phi,i)\\ \nonumber.
\end{displaymath}
%
In the last expression we assume that $S(b,\phi)$ is constant. 
This corresponds to $I(\mu) \propto \mu$ at the surface, where $\mu$
denotes the cosine of the angle of the propagation direction of the
outgoing photon and the normal to the spheres surface: a fixed $db$
corresponds to a physical length $ds = db/\mu$ on the sphere.  
If $I(\mu)$ were constant, this would imply that the sphere should
appear brighter per unit $b$. 
A constant surface brightness profile requires the directional
dependence for $I(\mu) \propto \mu$ to correct for this. 
Indeed, this is what is expected for the escape of Ly$\alpha$ photons
from static, extremely opaque media (see \cite{Ahn01}; their
Fig~4 and accompanying discussion).

It is worth stressing that the analytic calculation is clearly not a
complete analytic calculation, and we do not expect perfect agreement:
we {\it assumed} a functional form for the surface brightness profile
[or for $I(\mu)$].

It is important to realise that $I(\mu)$ itself may
depend on the spectrum. I.e., analytic solutions exist for $J(x) =
\int_0^1 I(x,\mu) d\mu$ at the boundary, and {\it approximate}
expressions for $I(\mu) =\int dx I(x,\mu)$, but none for $I(x,\mu)$
itself.

Fig~\ref{fig:comparison}shows some examples of analytic vs. full MC
spectra.
The area under the spectra is the same in all cases. 
The left panel shows the results for different rotational velocitiies
in the case of $\tau_{H}=10^7$ and an observer located perpendicular
to the axis of rotation ($\mu=0$ in the notation of the main body of
the paper, $i=0$ in the scheme of Fig~\ref{fig:scheme}). 
The right panel shows the results for different viewing angles in the
case of $\tau_{H}=10^7$ and a rotational velocity of $V_{\rm
  max}=300$\kms.

The two methods clearly give good agreement, though not perfect. In
particular, the left panel shows that the MC gives rise to a spectrum that is
slightly more concentrated towards the line centre. 
It is easy to get better agreement my modifying the surface brightness
profile (e.g. if $S(b)=2-b$, then the analytic calculation agrees
better).
In any case, the analytic calculation closely captures the results
obtained from the full calculations from the MC simulations.  
As such, they are extremely useful and provide us with a quick tool to
verify our calculations at the first order level.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{vary_vel.pdf}
  \includegraphics[width=0.48\textwidth]{vary_angle.pdf}
\end{center}
\caption{
Comparison of the Monte Carlo results agains the analytic
solution. The left panel explores the results of different velocities.
The right panel presents the results for two different observers:
paralel and perpendicular to the rotationa axsi, $\mu=1$ and $\mu=0$
respectively.
\label{fig:comparison} }   
\end{figure*}
